---
title: "Inverse-Simpson"
author: "Natalie Graham"
date: "2022-12-21"
output: html_document
---
```{r, message=FALSE, warning=FALSE}
# data management
library(tidyverse)
library(readr)
library(ggplot2); theme_set(theme_bw())
library(here) # set relative file location

# to calculate beta diversity
library("vegan")

# color palettes
library(colorspace)
library(RColorBrewer)

# to run linear mixed effects model
library(lme4)
library(lmerTest)
library(performance) # R2 
library(sjPlot) # Corr matrix

```

```{r, message=FALSE, warning=FALSE}
# read in beating data and niclassify results
beating <- read_csv(here("data/beatingRedone.csv"))
predict <- read_csv(here("data/niclassify-predictions_replacedfalseendemics.csv"))

# read in phylogeny
tree <- read.tree(here("data/postcutoff.tre"))

```

# read in data from species delimitation using ptp; 
details at https://cme.h-its.org/exelixis/web/software/PTP/index.html

# read in csv with ptp results
```{r, message=FALSE}
ptpDir <- here("data/by_order")
ptpFiles <- list.files(ptpDir, pattern = ".csv")
species <- lapply(file.path(ptpDir, ptpFiles), read_csv)
species <- do.call(rbind, species)
```

# join data files together
```{r, message=FALSE, warning=FALSE}
beating <- beating %>%
  left_join(predict) %>% 
  left_join(species) 

beating <- beating %>% 
  unite("delim_spp", c(orderGrp, delim), sep = "_", remove = FALSE)

wo_beating <- beating %>%
    filter(!site == "kohalaYng" & 
         !site == "LSAG" &
         !site == "hippnet") %>%
    # filter out the plots with more than one sd from mean of sampling time
  filter(!sitePlot == "flow1973_110" & 
         !sitePlot == "flow1973_107" &
         !sitePlot == "kohalaOld_05a" &
         !sitePlot == "waikamoi_06" &
         !sitePlot == "kamakou_15" &
         !sitePlot == "kokee_21a") %>%
  arrange(flowAge) %>%
  mutate(volcano = factor(volcano, unique(volcano))) %>%
  mutate(site = factor(site, unique(site))) %>%
  mutate(sitePlot = factor(sitePlot, unique(sitePlot)))

# filter the non-native taxa
endemic <- wo_beating %>%
  filter(predict == "endemic")
```

# Hypothesis 1 - Species richness increases with community age
[ x ] Prepare data for plot level correlations - *not including order as random effect*
```{r}

# First summarise data for number of OTUs per plot and number of species per plot
plotlevel <- endemic %>%
  group_by(volcano, site, flowAge, sitePlot) %>%
  summarise(`Num. OTU` = n_distinct(otu),
         `Num. Species` = n_distinct(delim_spp),
         `Num. Plants` = n_distinct(plant)) 

# Next summarise data for inverse simpson calculation at plot level
# First make a matrix at plot level (columns = 2737 OTUs, rows = 50 sampling plots) with read abundances

plotAbun <- endemic %>% 
  group_by(sitePlot, otu) %>%
  summarise(abundance = sum(abundance)) %>%
  ungroup() %>%
  pivot_wider(names_from = otu, values_from = abundance) 

plotAbunMatrx <- plotAbun %>%
  mutate_if(is.numeric, replace_na, replace = 0)
plotAbunMatrx <- plotAbunMatrx %>%
  select(-sitePlot)
plotAbunMatrx <- as.matrix(as.data.frame(plotAbunMatrx))

# Calculate inverse simpson based on raw abun of each otu at each site
invSimpPlot <- diversity(plotAbunMatrx, "inv")

invSimpValues <- as.data.frame(cbind(sitePlot = as.character(plotAbun$sitePlot), invSimpPlot))

invSimpValues$invSimpPlot <- as.numeric(invSimpValues$invSimpPlot)

# add back in site info
plotlevel <- plotlevel %>%
  left_join(invSimpValues) %>%
  rename(`Inv. Simp.` = invSimpPlot)

# Turn raw number of reads into relative abundance by dividing each value by
# sample total abundance
comm <- decostand(plotAbunMatrx, method = "total")

# Calculate inverse simpson based on the relative abun of each otu at each site
invSimpREl <- diversity(comm, "inv")

# No change in inv simp values
invSimpREl == invSimpPlot
```

[x ] Plot pattern of spp, otu, inv simp, plant by time -- don't include order *Fig. 2 Panel A*
```{r, message=FALSE}
# drop the 50 levels so ggpairs will run a corr matrix
plotlevel2 <- plotlevel %>%
  ungroup() %>%
  mutate(`Ln Flow Age` = log(flowAge)) %>%
  select(-c(flowAge, volcano, site, sitePlot)) 


plotlevel2 %>%
  pivot_longer(cols = 
                 c("Num. OTU", 
                   "Num. Species", 
                   "Num. Plants", 
                   "Inv. Simp."), 
               names_to= "Richness") %>%
  mutate(Richness = factor(Richness, 
                           levels=c("Num. OTU",
                                    "Num. Species",
                                    "Inv. Simp.",
                                    "Num. Plants"))) %>%
  ggplot(aes(x = `Ln Flow Age`, y = value, color = Richness)) +
  geom_point(cex = 3) +
  facet_wrap(~Richness, 
             scale = "free", ncol = 2) +
  scale_color_manual(values = c("#d7191c","#fdae61","#a6d96a","#1a9641")) +
  ylab("")+
  theme(legend.title= element_blank(),
        legend.position = "") +
  xlab("Ln Substrate Age (yrs)")

#ggsave("Richness_ms/figures/diversity-time5.jpg", width = 3, height = 3.5, units = "in")
    
```

[ x] Show the correlation for these richness measures and time, w/o order as a random effect

```{r, message=FALSE}
# correlation matrix
ggpairs(plotlevel2)
```

# SEM to consider effect of plant diversity increasing relative to ln substrate age increasing on the dependant variable OTU richness 
[ ] SEM for OTU richness
[ ] SEM for Inv. Simp.
- Question for Jun - We test SEM with df that includes the smaller orders but we exclude those smaller orders for our mixed effect model bc they don't have enough coverage, does this make sense in the context of the SEM being agnostic to order but the multivariate analysis taking it into consideration? I think so but good to have another's opinion

```{r}
plotlevel
```

# Hypothesis 1 (continued) - Species richness increases with community age
[x] Prepare data for linear mixed effects model *now with arthropod order as a random effect* 
[x] Calculate inv simp at order level for the orders that are large enough
```{r, message=FALSE}

# first need matrix at plot level with arthropod included
plotAbun <- endemic %>% 
  group_by(sitePlot, otu, orderGrp) %>% #ordergroup
  summarise(abundance = sum(abundance)) %>%
  ungroup() %>%
  pivot_wider(names_from = otu, values_from = abundance) %>%
  unite("orderPlot", c(sitePlot, orderGrp), remove = TRUE)

plotAbunMatrx <- plotAbun %>%
  mutate_if(is.numeric, replace_na, replace = 0)
plotAbunMatrx <- plotAbunMatrx %>%
  select(-orderPlot)
plotAbunMatrx <- as.matrix(as.data.frame(plotAbunMatrx))

# Calculate inverse simpson based on abun of each otu at each plot
invSimpPlotOrder <- diversity(plotAbunMatrx, "inv")

# Bind data from inv simp calculatoion to the dataframe with abundance
invSimpPlotOrder <- as.data.frame(cbind(orderPlot = plotAbun$orderPlot, invSimpPlotOrder))

invSimpPlotOrder$invSimpPlotOrder <- as.numeric(invSimpPlotOrder$invSimpPlotOrder)

# make df with richness summary and site info
#sitePlotinfo <- endemic %>%
  #select(volcano, site, sitePlot, flowAge) %>% distinct()

# make df with richness summary and site info
richotuplot <- endemic %>%
  group_by(orderGrp, site, sitePlot, flowAge, volcano) %>% 
   summarise(n_species = n_distinct(delim_spp),
             n_otu = n_distinct(otu)) 

invSimpPlotOrder <- invSimpPlotOrder %>%
  separate(orderPlot, c("site", "sitePlot", "orderGrp"), sep = "_") %>%
  unite("sitePlot", c(site, sitePlot), remove = TRUE) %>%
  left_join(richotuplot) 
invSimpPlotOrder
```

Plot the relationship of community age to species richness (no. otus & no. spp) and diversity (inv simpson)
```{r}
invSimpPlotOrder %>%
  pivot_longer(cols = c(n_otu, n_species, invSimpPlotOrder), names_to = "richness") %>%
  ggplot(aes(x = log(flowAge), y = value, group = richness)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), size = 1, se = TRUE, aes(colour = richness)) +
  facet_wrap(~orderGrp, scale = "free")

```

[x] Remove the arthropod orders that are undersampled by vegetation beat sampling in the understory. 
- Cutoff is < 100 OTUs - based on the spec accum curves 
[x] Add in summary of plants
```{r}

invSimpPlotOrder %>%
  group_by(orderGrp) %>%
  summarise(nOTU = sum(n_otu)) %>%
  arrange(nOTU)

# summarise the number of plants that were sampled within each plot, per order group
uniquePlant <- wo_beating %>%
  group_by(volcano, site, flowAge, sitePlot) %>% 
  summarise(n_plant_sampled = n_distinct(plant)) %>%
  arrange(desc(n_plant_sampled))

Ocorr <- invSimpPlotOrder %>%
  left_join(uniquePlant) %>% # join the number of plants
  # subset by orders with more than 100 OTUS
  filter(!orderGrp == "Hymenoptera" & 
         !orderGrp == "Myriapoda" &
         !orderGrp == "Neuroptera" &
         !orderGrp == "Malacostraca") %>%
  mutate(`ln FlowAge` = log(flowAge)) %>%
  rename(`Num. OTU` = n_otu,
         `Num. Species` = n_species,
         `Inver. Simp.` = invSimpPlotOrder,
         `Num. Plants` = n_plant_sampled) 

Ocorr # this is now the object for mixed effect model
```

[x ] Mk correlation matrix for each arthropod order
```{r, message=FALSE}

# define a function to plot for each order
pairsFunc <- function(ordernm, title){
Ocorr <- Ocorr %>%
  select(-c("sitePlot", 
            "flowAge",
            "site",
            "volcano"
            )) %>%
  filter(orderGrp == ordernm) %>%
  select(-orderGrp)
ggpairs(Ocorr) +
  ggtitle(paste0(ordernm))
}

# vector of order names to pass to function
ordernm=unique(invSimpPlotOrder_lgr$orderGrp)
# use list apply to store each of the plots in a list
listplots <- lapply(ordernm, pairsFunc, title)
names(listplots) = ordernm # name each

# iterate over the list of plots and save each by name
for(i in 1:length(ordernm)){
  ggsave(plot = listplots[[i]], 
         file = paste(names(listplots)[i],".jpg",sep=""))
}

```

